Investigating Neolithic pottery usage from Guijiabao in the Yanyuan Basin in Southwest China using organic residue analysis

### 2023-01-31 analysis/paper/paper.Rmd
working on the discussion section

<!--info about Gaoshan-->
Gaoshan site is located in the southwestern margin of the Chengdu Plain, an alluvial plain between the Sichuan Basin and the Tibetan Plateau (Figure \@ref(fig:SW-china-map)). Excavations at Gaoshan revealed abundant pottery, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015]. According to the pottery forms and decorations, Gaoshan can be categorized into the earliest phase of the Baodun period, the earliest Neolithic culture (ca. 4700-3700 BP) in the Chengdu Plain [@Guedes2013]. It is proposed that Gaoshan could be even older than the Baodun culture according to a burial context, which was found in the stratigraphic layer earlier than the Baodun period [@Gaoshan2017]. A large quantity of carbonized rice grains was found at Gaoshan with a relatively small number of foxtail/broomcorn millets, suggesting rice is the main crop supplemented by millets [@JiangGaoshan2017]. This also indicates the spread of rice cultivation from the Middle and Lower Yangzi valley to Sichuan in southwest China occurred at least 4500 years ago [@Chi2010]. Moreover, recent research on animal bones in 2015 and 2016 shows that pigs and dogs were already domesticated, of which pigs are the main domestic animal [@He2020]. The primary subsistence strategy at Gaoshan could be rice/millet cultivation with domesticated pigs.

@@

The results show that most potsherd samples from Guijiabao contain fatty acids from ruminant adipose fats with one potsherd (SYG-TN13-E23-3) suggesting an origin of non-ruminant adipose fat (Figure \@ref(fig:isotopic-plots): A). The potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well. A few number of potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003; @Dudd1998; @Suryanarayan2021]. Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020]. In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with the δ^13^C values of C3/terrestrial plants, suggesting the ruminant animals were likely fed on C3 plants. There is only one potsherd (SYG-TN13-E22-2#2) have a relatively enriched δ^13^C value, indicative of a C4 plant diet for the ruminant animal.

@@

# Discussion
<!--A brief summary of interesting results-->
Our results show that food categories indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were mainly dominated by terrestrial animals, particularly ruminant fats or mixtures of terrestrial animals and plants. Moreover, it is striking to identify a non-ruminant source from a Guijiabao potshard that might originate from terrestrial plants. Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support that those ceramics were used as cooking or storage pots. The low yield might be related to the part of pottery samples analyzed in this study are mostly base fragments [@Evershed2008experimental; @Kwak2020]. But we also found a base sample dating to ca. 4500 BP. contains a rich composition of long-chain fatty acids. We recognize the limitation for interpreting food choices over time or differences across regions because of our small sample sizes. However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices or the local subsistence of Guijiabao.

<!--discussion about possible plant origins-->
One aim of our study is to explore whether new crops, such as C3 plants (eg. rice), C4 plants (eg. millet), or new species (eg. pigs/horses) became an important part of local diet in Yanyuan during the Neolithic transition. We successfully identified non-ruminant fats from a Guijiabao potshard that might be from terrestrial plant/C3 plants based on similar isotopic values reported by other studies [@Shoda2018; @Kwak2020]. Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for accessing food categories. Also, recent macro-botanical and phytolith studies at Guijiabao suggest the presence of a mixing farming of rice and millet, indicating that the origin of C3 plants from our sample could come from rice [@Hao2022 @Huan2022]. It is also possible other grasses classified as C3 plants, such as Pennisetum, were part of the local diet [@Guijiabao2016]. We should note that freshwater fish could also be the origin of non-ruminant fats since C3 plants and freshwater fish have similar isotopic values [@Lucquin2016]. Fish bones and net sinkers were also found at Guijiabao which suggests the possibility of fishing. There are no C4 plants, such as millets, directly detected in our sample, which contradicts the predominance of millet grains found at Guijiabao. But it is hard to make an inference about the role of rice/millet because the results might be caused by small sample effects or sampling strategy.

<!--discussion about possible ruminant fat origin-->
Ruminant fats in the Neolithic potsherds around ca. 4500 BP. from Guijiabao and Gaoshan could be derived from deer or horse according to unearthed faunal remains from the sites, with stone tools associated with hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a]. But whether horses were part of local food choices in the Neolithic period is questionable. First, the Pleistocene and Neolithic sites with wild horse bones primarily concentrated in northern China [@Yuan2006]. Second, domesticated horses first found in southwest China were much later from the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004]. The evidence from Guijuabao also confirms a later presence of horse remains. In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value located in ranges of deer or sheep fats based on a study from England [@Evershed2002].  It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were part of food sources during this period [@Zhang2016]. A detailed analysis of faunal remains from Guijiabao is still in progress, allowing further evaluation of our findings.

In general, our results suggest that Guijiabao residents relied mostly on ruminants fats with terrestrial plants as supplements. The overall depleted δ^13^C value shows that the diet of ruminants was mainly C3 plants that could come from wild weeds [@Guijiabao2016].

--Gaushan-90 pig, plant-most are rice: 95% and a few millet  
--ecological zones

A large-scale analysis of ceramic across the region may offer insight into regional variations in food choices and the spread of agriculture. Our next step is to increase the sample size to include pots from different time periods and different types of pottery to identify any variation through time and also try different methods, such as phytolith analysis to further support our statement about rice cultivation.

>>>>

<!--info about Gaoshan-->
Gaoshan site is located in the southwestern margin of the Chengdu Plain, an alluvial plain between the Sichuan Basin and the Tibetan Plateau (Figure \@ref(fig:SW-china-map)). Excavations at Gaoshan revealed abundant pottery, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015]. According to the pottery forms and decorations, Gaoshan can be categorized into the earliest phase of the Baodun period, the earliest Neolithic culture (ca. 4700-3700 BP) in the Chengdu Plain [@Guedes2013]. It is proposed that Gaoshan could be even older than the Baodun culture according to a burial context, which was found in the stratigraphic layer earlier than the Baodun period [@Gaoshan2017]. A large quantity of carbonized rice grains was found at Gaoshan with a relatively small number of foxtail/broomcorn millets, suggesting rice is the main crop supplemented by millets [@JiangGaoshan2017]. This also indicates the spread of rice cultivation from the Middle and Lower Yangzi valley to Sichuan in southwest China occurred at least 4500 years ago [@Chi2010]. Moreover, recent research on animal bones in 2015 and 2016 shows that pigs and dogs were already domesticated, of which pig is the dominant animal [@He2020]. The primary subsistence strategy at Gaoshan could be rice cultivation with domesticated pigs, supplemented by millet cultivation.

@@

The results show that most potsherd samples from Guijiabao contain fatty acids from ruminant adipose fats with one potsherd (SYG-TN13-E23-3) suggesting an origin of non-ruminant adipose fat (Figure \@ref(fig:isotopic-plots): A). The potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well. A few number of potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003; @Dudd1998; @Suryanarayan2021]. Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020]. Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for accessing food categories.

In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with the δ^13^C values of C3/terrestrial plants, suggesting the ruminant animals were likely fed on C3 plants. There is only one potsherd (SYG-TN13-E22-2#2) have a relatively enriched δ^13^C value, indicative of a C4 plant diet for the ruminant animal.

@@

# Discussion
<!--A brief summary of interesting results-->
Our results show that food choices indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were mainly dominated by ruminant fats during the Neolithic period. Moreover, it is striking to identify a non-ruminant source from a Guijiabao potshard that might originate from C3 plants. Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support that those ceramics were used as cooking or storage pots. The low yield might be related to the part of pottery samples analyzed in this study are mostly base fragments [@Evershed2008experimental; @Kwak2020]. But we also found a base sample dating to ca. 4500 BP. contains a rich composition of long-chain fatty acids, reflecting an excellent preservation. We recognize the limitation for interpreting food choices over time or differences across regions because of our small sample sizes. However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices or the local subsistence at Guijiabao.

<!--discussion about possible plant origins-->
One aim of our study is to explore whether new crops, such as C3 plants (eg. rice), C4 plants (eg. millets), or new species (eg. pigs/horses) became an essential part of local diet in Yanyuan during the Neolithic transition. We successfully identified non-ruminant fats from a Guijiabao potshard that might be from terrestrial plant/C3 plants based on similar isotopic values reported by other studies [@Shoda2018; @Kwak2020]. It is also possible other native grasses classified as C3 plants, such as pennisetum and chenopodiaceae, were part of the local diet [@Guijiabao2016]. Surprisingly, we only observed signals of C3 plants, despite the fact that C4 plants (foxtail and broomcorn millets) are the major crop based on the recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022]. Due to our small samples, it is hard to determine the role of rice/millet during the Neolithic period. Although freshwater fish could be another origin of non-ruminant fats that can suggested by the finding of fish bones and net sinkers at Guijiabao [@Lucquin2016; @Guijiabao2017b], we did not detect any presence of biomarkers for aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid from our potsherds [@Evershed2008marine; @Hansel2004; @Kwak2017].

<!--discussion about possible ruminant fat origin-->
Ruminant fats identified in the Neolithic potsherds around ca. 4500 BP. from Guijiabao and Gaoshan could be derived from deer or horse according to unearthed faunal remains from the sites, accompanied with stone tools for hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a]. But whether horse was part of local food choices in the Neolithic period is questionable. First, the Pleistocene and Neolithic sites with wild horse bones primarily concentrated in northern China [@Yuan2006]. Second, domesticated horses first found in southwest China were much later from the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004]. The evidence from Guijuabao also confirms a later presence of horse remains. In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value consistent with those values of deer and sheep fats based on global references [@Evershed2002]. It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were part of food sources during this period [@Zhang2016]. A detailed analysis of faunal remains from Guijiabao is still in progress that can provide further information.

We observed a mixture of ruminant fats and terrestrial plants based on a number of Guijiabao potsherds dating to the historical period (ca. 1000 BC.). However, we are unable to determine whether there is a shift in diet over time due to our limited sample size. But the overall depleted δ^13^C values for most potsherds suggest C3 plants might be an important diet for local animals and could also be included in the local diet in Yanyuan during the Neolithic transition. This corresponds with the timing when new crops spread into southwest China. Despite of no C4 plant origin in our samples, one potshard implies a source of C4 plant-fed ruminant. In general, the food source detected from our samples comprised mainly of ruminant fats that are more likely from wildlife. This might be related to local ecological zones and environments where wild animals were easier to obtain. Since zooarchaeological analysis and other studies on stone tools are still in progress, we could only rely on preliminary excavation reports with limited details about associated artifacts. A large-scale organic residue analysis of ceramics across different pottery forms in a wider region may offer insight into temporal and regional variations in food choices during the Neolithic transition.
